Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-05 analysis/paper/paper.Rmd
end of day commit.

## Aspatial performance

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). All road network projection method variants perform better for all metrics than the "cutblocks only" method where cutblocks were the only disturbance considered. All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the outside cutblock road disturbance footprint more than low density methods, and underproject the outside cutblock distance to roads, but all method variants give reasonably good projections of the outside cutblock distance to roads. All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

@@

## Spatial performance

Difference maps highlight several potential causes of discrepancy between our projections and the observed road network. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a). However, the cost raster is not sufficiently detailed to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:proj-map-fig)c). All these discrepancies could be addressed with more accurate harvest and cost information.

<!--# TODO: I don't know that the switchback issue could be addressed with more accurate cost information. Wouldn't you need to account for the direction of the slope and cost according to that in order to induce that behaviour? I guess that would be better cost info but is more complicated than this implies.  -->

@@

As expected, none of the projection methods can accurately project road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The performance of the "cutblocks only" method could not be assessed for road presence or disturbance footprint since no roads were projected. However, the "cutblocks only" method performed better than any other method for the forestry disturbance footprint. This reflected the fact that in the Revelstoke TSA most new roads were built inside  cutblocks, so the number of false negatives caused by not projecting roads was small. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. All of these could be improved with more detailed high resolution cost & harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results may be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->

>>>>

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). All road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered. Within all cutblocks, the Hardy method most accurately projects road density, and least accurately projects distance to road. Low density method variants most severely underproject road density within cutblocks, and most accurately project distance to road within cutblocks. These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method more accurately projects ... All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy. Another source of discrepancy between observed and projected networks in this mountainous landscape are switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c). None of the simple algorithms considered here are capable of projecting switchbacks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.

As expected, none of the projection methods yield accurate spatial projections of road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`). Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

@@

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. stream crossing behaviour could improved with more detailed high resolution cost. harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. Compare to Hardy results in eastern canadian shield - better road density projections in that landscape suggests better performance of simple iterative least cost path algorthms in more moderate terrain. 
- Recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->
